Reasoning about a Capability Machine with Local Capabilities

### 2017-06-16 popl-2018/paper.tex
finished first read through of LR section - still contemplating whether the last part about execute condition should be updated to talk about an adversary

% Uncomment me to disable comments
\renewcommand\lau[1]{}
\renewcommand\dominique[1]{}
\renewcommand\lars[1]{}

@@

The \emph{observation relation} $\observations$ defines configurations that we
consider safe. Roughly speaking, a configuration is safe if whenever it halts,
there is a private future world that the resulting memory satisfies. Notice that
failing is considered safe behavior. In fact, the machine often resorts to
failing when an unauthorized access is attempted, like loading from a capability
without read permission. This is similar to \citet{Devriese:2016ObjCap}'s logical
relation for an untyped language, but unlike typical logical relations
for typed languages, which require that programs do not fail.

The \emph{register-file relation} $\stdrr$ defines safe register-files as those
that contain safe words (i.e.\ words in $\stdvr$) in all registers but $\pcreg$.
The \emph{expression relation} $\stder$ defines that a word is safe to use as a
program counter if it can be plugged into a safe register file (i.e.\ a register
file in $\stdrr$) and paired with a memory satisfying the world to produce a
safe configuration.

The \emph{value relation} $\stdvr$ defines when words are safe, i.e.\ when they
cannot be used to break the protocols defined by the world. Non-capability data
is always safe because it provides no authority. Capabilities give the authority
to manipulate memory and potentially break memory protocols, so they need to
satisfy certain conditions to be safe. In \figurename~\ref{fig:perm-conds}, we
define such a condition for each kind of permission a capability can have (we
will call these permission-based conditions).

For capabilities with read permission, the $\readCond{}$ ensures that it can
only be used to read safe words, by requiring that the region governing the
relevant part of memory, guarantees that it contains only words in the value
relation. Generally speaking, global capabilities should only depend on
permanent regions. Therefore we use the function $\var{localityReg}(\gl,W)$,
which projects out all active regions when $\gl$ (the locality) is local, but
only the permanent regions when $\gl$ is global. The read condition is defined
in terms of a standard region $\iota^\pwl_{\start,\addrend}$ which requires all
the words in the range $[\start,\addrend]$ to be safe.

For a capability with write permission, $\writeCond{}$ must be
satisfied for the capability's range of authority. Such a capability
may be used to write any word, so the region governing the relevant
memory must (at least) allow any safe word to be stored. More
precisely, for write-local capabilities it needs to allow arbitrary
safe values to be stored, while for write capabilities, storing values
that are both non-local and safe is sufficient.  This is done in the
value relation by the regions the write condition is applied with. For
write-local capabilities $\iota^\pwl$ is used, and for write
capabilities we use another standard region
$\iota^\nwl_{\start,\addrend}$ which requires that the words in
$[\start,\addrend]$ are non-local and safe.  Like before, we use
$\var{localityReg}$ to pick an appropriate region for the capability's
locality. The condition for the region to be address-stratified is a
technical requirement, defined in the technical appendix.
An interesting property of the read and write condition is that they
prohibit global write-local capabilities which, as discussed in
Section~\ref{sec:stack-and-return-pointer}, is necessary for the safe
use of local capabilities.

>>>>

% Uncomment me to disable comments
% \renewcommand\lau[1]{}
% \renewcommand\dominique[1]{}
% \renewcommand\lars[1]{}

@@

The \emph{observation relation} $\observations$ defines what configurations we
consider safe. As previously states, a configuration is safe with respect to a
world when the execution of said configuration does not break the memory
protocols of the world. Roughly speaking, this means that when the execution of
a configuration halts, then there is a private future world that the resulting
memory satisfies.  Notice that failing is considered safe behavior. In fact, the
machine often resorts to failing when an unauthorized access is attempted, like
loading from a capability without read permission. This is similar to
\citet{Devriese:2016ObjCap}'s logical relation for an untyped language, but
unlike typical logical relations for typed languages, which require that
programs do not fail.

The \emph{register-file relation} $\stdrr$ defines safe register-files as those
that contain safe words (i.e.\ words in $\stdvr$) in all registers but $\pcreg$.
The \emph{expression relation} $\stder$ defines that a word is safe to use as a
program counter if it can be plugged into a safe register file (i.e.\ a register
file in $\stdrr$) and paired with a memory satisfying the world to become a
safe configuration. This means that integers and non-executable capabilities
(e.g. readonly and enter capabilities) are considered safe as program counters
because when they are plugged into a register file and paired with a memory, the
execution will immediately fail, which is safe.

The \emph{value relation} $\stdvr$ defines when words are safe, i.e.\ when they
cannot be used to break the memory protocols defined by the world. We make the
value relation as liberal as possible by considering what is the most we can
allow an adversary to use a capability for without breaking the memory
protocols. Non-capability data is always safe because it provides no
authority. Capabilities give the authority to manipulate memory and potentially
break memory protocols, so they need to satisfy certain conditions to be
safe. In \figurename~\ref{fig:perm-conds}, we define such a condition for each
kind of permission a capability can have (we will call these permission-based
conditions).

For capabilities with read permission, the $\readCond{}$ ensures that it can
only be used to read safe words by putting an upper bound on what can be
required by the region governing the memory the read capability gives access
to. An adversary can read anything the capability governs, so all the words they
can possibly read must be safe - that is they must be in the value relation. The
read condition also depends on the locality of the capability because, generally
speaking, global capabilities should only depend on permanent
regions. Concretely to make sure global capabilities only depend on permanent
regions, we use the function $\var{localityReg}(\gl,W)$, which projects out all
active regions when $\gl$ (the locality) is local, but only the permanent
regions when $\gl$ is global. The upper bound on the region is enforced in terms
of a standard region $\iota^\pwl_{\start,\addrend}$ which requires all the words
in the range $[\start,\addrend]$ to be safe.  The region in the world that
actually governs the part of memory the capability gives access to must as most
be as restrictive as the standard region.  \lau{Here I would have liked to point
  back to the bit about step-indexing in the intuition section, but
  unfortunately the ``step down'' happens in the standard region which we have
  not included.}

For a capability with write permission, $\writeCond{}$ must be satisfied for the
capability's range of authority. An adversary can use such a capability to write
any word they can get a hold of, and we can safely assume that they can get a
hold of any safe word, so the region governing the relevant memory must allow
any safe word to reside there. In the name of making the logical relation as
liberal as possible, we make this a lower bound of what the region may allow.
\lau{Maybe add: Conceptually, the lower bound also conveys the amount of faith
  one can put in the contents of a piece of memory that an adversary has been
  given write access to.}  For write capabilities, we also have to take into
account that we have two flavours of write permissions namely write and
write-local. In the case of write-local capabilities, the region needs to at
least require any safe word to be written, but in the case of write
capabilities, the capability cannot be used to write local capabilties, so the
region needs only to at least require safe non-local values.  In the write
condition, this is handled by parameterising it with a region. For the
write-local capabilities the write condition is applied with the standard region
$\iota^\pwl$ and for the write capabilities we use a different standard region
namely $\iota^\nwl_{\start,\addrend}$ which requires that the words in
$[\start,\addrend]$ are non-local and safe. Like before, we use
$\var{localityReg}$ to pick an appropriate region based on the capability's
locality. Finally the region is required to be address-stratified. Intuitively,
this means that if a region accepts two memory segments, then it must also
accept every memory segment ``inbetween'' that is every memory segment where
each address contains a value from one of the two memory segments. An
interesting property of the read and write condition is that they prohibit
global write-local capabilities which, as discussed in
Section~\ref{sec:stack-and-return-pointer}, is necessary for the safe use of
local capabilities.
